---
title: "DifferentialAbundance"
author: "JM"
date: "2025-03-01"
output: html_document
---


### Use corncob to look for differentially abundant taxa by the continuous variable of total tissue loss

```{r, echo=FALSE}
library(phyloseq)
library(corncob)
library(plyr)
library(dplyr)
library(reshape2)
library(tibble)
library(ggplot2)
library(ggbreak)
```

# What microorganisms are significantly changing with changes in tissue loss? Use Jan samples only; total tissue loss in January

Data prep:
# use an ASV table that has been filtered to remove low-abundance ASVs, no blanks, replaced sequences with ASV numbers in both otu table and taxa table, added a column named "Species" to the taxa table to fill in later with ASV number

```{r, echo=FALSE}
otu <- read.table("silva_nochloronomito_otu_table_ps5_BS2_jan_corncob.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_nochloronomito_taxa_table_ps5_BS2_jan_corncob.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata_ps5_BS2_jan.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps5_BS2_jan <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
                        sample_data(samples), 
                        tax_table(taxon))
ps5_BS2_jan  #756 taxa and 63 samples

ps_corn = subset_samples(ps5_BS2_jan, Proportion_TissueLost_Jan21 != "nd")
ps_corn #756 taxa and 54 samples
ps_corn <- prune_taxa(taxa_sums(ps_corn) > 1, ps_corn) #remove taxa not found in remaining samples
ps_corn #651 taxa and 54 samples, removed 105 taxa only found in samples without tissue loss data

set.seed(1)
tissueJan.da <- differentialTest(formula = ~ Proportion_TissueLost_Jan21, 
                                 phi.formula = ~ Proportion_TissueLost_Jan21,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Proportion_TissueLost_Jan21,
                                 test = "LRT", boot = FALSE,
                                 data = ps_corn,
                                 fdr_cutoff = 0.05)

# All p-values are NA! Model overspecified!

```


### Use corncob to look for differentially abundant taxa by category: treatment type

####### JANUARY
# use an ASV table that has been filtered to remove low-abundance ASVs, no blanks, replaced sequences with ASV numbers in both otu table and taxa table, added a column named "Species" to the taxa table to fill in later with ASV number

```{r, echo=FALSE}
otu <- read.table("silva_nochloronomito_otu_table_ps5_BS2_jan_corncob.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_nochloronomito_taxa_table_ps5_BS2_jan_corncob.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata_ps5_BS2_jan.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps5_BS2_jan <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
                        sample_data(samples), 
                        tax_table(taxon))
ps5_BS2_jan  #756 taxa and 63 samples


# What microorganisms are significantly different by treatment type? 
# DA determined relative to first reference level, so set this manually to the "none" treatment
sample_data(ps5_BS2_jan)$Treatment<-factor(sample_data(ps5_BS2_jan)$Treatment,levels=c("none","probiotic bag","probiotic paste","control bag","control paste"))

set.seed(1)
treatment.da <- differentialTest(formula = ~ Treatment, 
                                 phi.formula = ~ Treatment,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Treatment,
                                 test = "Wald", boot = FALSE,
                                 data = ps5_BS2_jan,
                                 fdr_cutoff = 0.05)

summary(treatment.da)
treatment.da
treatment.da$significant_taxa 
# No differentially abundant taxa


#### See if there are any DA families
# collapse ASVs into higher taxonomic groups for DA analysis
jan_family <- ps5_BS2_jan %>% phyloseq::tax_glom("Family")
jan_family # 183 taxa (families) and 63 samples


# DA determined relative to first reference level, so set this manually to the "none" treatment
sample_data(jan_family)$Treatment<-factor(sample_data(jan_family)$Treatment,levels=c("none","probiotic bag","probiotic paste","control bag","control paste"))

set.seed(1)
treatment.da <- differentialTest(formula = ~ Treatment, 
                                 phi.formula = ~ Treatment,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Treatment,
                                 test = "Wald", boot = FALSE,
                                 data = jan_family,
                                 fdr_cutoff = 0.05)

summary(treatment.da)
treatment.da
treatment.da$significant_taxa # 2 differentially abundant families

#quick look at results with corncob's plotting feature
plot(treatment.da, c("Family"))
# Positive values mean that taxon is enriched in the treatment in the header relative to no treatment
# DA families: Comamonadaceae, Micrococcaceae

# create a function that will return the values of interest

sigtaxto_df <- function(treatment.da, jan_family) {
  df.out <- c()
  for (i in 1:length(treatment.da$significant_models)) { #from 1 to the number of significant taxa
    df.out = rbind(df.out, treatment.da$significant_models[[i]]$coefficients[2, 1:4]) #bind the significant taxa with coefficient, std. error, t value, Pr. coefficients[2,] is the DA model output.
  }
  asv = as.vector(treatment.da$significant_taxa) #get significant families
  Class = as.vector(otu_to_taxonomy(data = jan_family, treatment.da$significant_taxa, level = c("Class")))
  Order = as.vector(otu_to_taxonomy(data = jan_family, treatment.da$significant_taxa, level = c("Order")))
  Family = as.vector(otu_to_taxonomy(data = jan_family, treatment.da$significant_taxa, level = c("Family")))
  
  df.bind <- as.data.frame(cbind(df.out, asv, Class, Order, Family))
  colnames(df.bind) <- c("Estimate", "StdError", "tvalue", "Pr", "asv", "Class", "Order", "Family")
  df.bind$Estimate = as.numeric(as.character(df.bind$Estimate)) #estimate column should be numeric 
  df.bind$StdError = as.numeric(as.character(df.bind$StdError)) #stderror column should be numeric 
  
  return(df.bind)
}

df <- sigtaxto_df(treatment.da, jan_family) #make a df of the corncob output and the taxonomy

### Save these data so you don't have to re-run the model ###
write.table(df, "DA-Families_treatment-Jan.txt", sep="\t",row.names = FALSE)


# retrieve relative abundances of significant taxa for plotting
jan_family #183 taxa and 80 samples
jan_family_ra<-transform_sample_counts(jan_family, function(OTU) OTU/sum(OTU))
jan_family_ra #183 taxa and 80 samples
otu_da = as(otu_table(jan_family_ra), "matrix")
taxon_da = as(tax_table(jan_family_ra), "matrix")
da<-t(otu_da)
da <- as.data.frame(da)
da <- tibble::rownames_to_column(da, "ASV")
taxon_da <- as.data.frame(taxon_da)
taxon_da <- tibble::rownames_to_column(taxon_da, "ASV")
taxon_da$Species<-NULL
taxon_da$Genus<-NULL
da_tax <- merge(taxon_da, da, by="ASV", all=FALSE)

# now to retrieve only the taxa that I want to plot
DA1 <- da_tax %>% filter(Family =='Comamonadaceae')
DA2 <- da_tax %>% filter(Family =='Micrococcaceae')
DA_jan <-rbind(DA1,DA2)

# Merge with metadata for plotting aesthetics
samples <- as.data.frame(samples)
samples <- tibble::rownames_to_column(samples, "sample")
DA_long<-melt(DA_jan,value.name="proportion",variable.name="sample",id.vars=c("ASV","Kingdom","Phylum","Class","Order","Family"))

DAplot <- merge(DA_long,samples,by="sample", all=TRUE)
# reduce decimal places in proporation column
DAplot$proportion<-round(DAplot$proportion,3)

cols<-c("none"="#000000","probiotic paste"="#FFC080","probiotic bag"="#FF8000","control paste"="#90BFF9","control bag"="#0000C0")
DAplot$Treatment<-factor(DAplot$Treatment,levels=c("probiotic bag","probiotic paste","control bag","control paste","none"))

pdf("DA-Families_Jan.pdf",bg ="white",width=8)
p2<-ggplot(DAplot,aes(x=Treatment,y=proportion,color=Treatment))+
  geom_boxplot()+
  geom_point()+
  facet_grid(Family~.)+
  theme_bw()+
  scale_color_manual(values=cols)
p2
dev.off()


```

### Use corncob to look for differentially abundant taxa by category: treatment type

####### OCTOBER
# use an ASV table that has been filtered to remove low-abundance ASVs, no blanks, replaced sequences with ASV numbers in both otu table and taxa table, added a column named "Species" to the taxa table to fill in later with ASV number


```{r, echo=FALSE}
otu <- read.table("silva_nochloronomito_otu_table_ps5_BS2_oct_corncob.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_nochloronomito_taxa_table_ps5_BS2_oct_corncob.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata_ps5_BS2_oct.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps5_BS2_oct <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
                        sample_data(samples), 
                        tax_table(taxon))
ps5_BS2_oct  #756 taxa and 80 samples


# What microorganisms are significantly different by treatment type?

# DA determined relative to first reference level, so set this manually to the "none" treatment
sample_data(ps5_BS2_oct)$Treatment<-factor(sample_data(ps5_BS2_oct)$Treatment,levels=c("none","probiotic bag","probiotic paste","control bag","control paste"))

set.seed(1)
treatment.da <- differentialTest(formula = ~ Treatment, 
                                 phi.formula = ~ Treatment,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Treatment,
                                 test = "Wald", boot = FALSE,
                                 data = ps5_BS2_oct,
                                 fdr_cutoff = 0.05)

summary(treatment.da)
treatment.da
treatment.da$significant_taxa #differentially abundant taxa
#ASV67 and ASV76

#quick look at results with corncob's plotting feature
plot(treatment.da, c("Family", "Genus"))
# Positive values mean that taxon is enriched in the treatment in the header relative to no treatment

# create a function that will return the values of interest

sigtaxto_df <- function(treatment.da, ps5_BS2_oct) {
  df.out <- c()
  for (i in 1:length(treatment.da$significant_models)) { #from 1 to the number of significant taxa
    df.out = rbind(df.out, treatment.da$significant_models[[i]]$coefficients[2, 1:4]) #bind the significant taxa with coefficient, std. error, t value, Pr. coefficients[2,] is the DA model output.
  }
  asv = as.vector(treatment.da$significant_taxa) #get significant ASVs
  Class = as.vector(otu_to_taxonomy(data = ps5_BS2_oct, treatment.da$significant_taxa, level = c("Class")))
  Order = as.vector(otu_to_taxonomy(data = ps5_BS2_oct, treatment.da$significant_taxa, level = c("Order")))
  Family = as.vector(otu_to_taxonomy(data = ps5_BS2_oct, treatment.da$significant_taxa, level = c("Family")))
  Genus = as.vector(otu_to_taxonomy(data = ps5_BS2_oct, treatment.da$significant_taxa, level = c("Genus")))
  
  df.bind <- as.data.frame(cbind(df.out, asv, Class, Order, Family, Genus))
  df.bind$asvgenus <- paste0(df.bind$Genus, " (", df.bind$asv, ")")
  colnames(df.bind) <- c("Estimate", "StdError", "tvalue", "Pr", "asv", "Class", "Order", "Family", "Genus", "asvtaxa")
  df.bind$Estimate = as.numeric(as.character(df.bind$Estimate)) #estimate column should be numeric 
  df.bind$StdError = as.numeric(as.character(df.bind$StdError)) #stderror column should be numeric 
  
  return(df.bind)
}

df <- sigtaxto_df(treatment.da, ps5_BS2_oct) #make a df of the corncob output and the taxonomy

### Save these data so you don't have to re-run the model ###
write.table(df, "DA_treatment-Oct.txt", sep="\t",row.names = FALSE)


# retrieve relative abundances of significant taxa for plotting
ps5_BS2_oct #756 taxa and 80 samples
ps5_BS2_oct_ra<-transform_sample_counts(ps5_BS2_oct, function(OTU) OTU/sum(OTU))
ps5_BS2_oct_ra #756 taxa and 80 samples
otu_da = as(otu_table(ps5_BS2_oct_ra), "matrix")
taxon_da = as(tax_table(ps5_BS2_oct_ra), "matrix")
da<-t(otu_da)
da <- as.data.frame(da)
da <- tibble::rownames_to_column(da, "ASV")
taxon_da <- as.data.frame(taxon_da)
taxon_da <- tibble::rownames_to_column(taxon_da, "ASV")
taxon_da$Species<-NULL
da_tax <- merge(taxon_da, da, by="ASV", all=FALSE)

# now to retrieve only the taxa that I want to plot
DA1 <- da_tax %>% filter(ASV =='ASV67')
DA2 <- da_tax %>% filter(ASV =='ASV76')
DA_oct <-rbind(DA1,DA2)

# Merge with metadata for plotting aesthetics
samples <- as.data.frame(samples)
samples <- tibble::rownames_to_column(samples, "sample")
DA_long<-melt(DA_oct,value.name="proportion",variable.name="sample",id.vars=c("ASV","Kingdom","Phylum","Class","Order","Family","Genus"))

DAplot <- merge(DA_long,samples,by="sample", all=TRUE)
# reduce decimal places in proporation column
DAplot$proportion<-round(DAplot$proportion,3)

cols<-c("none"="#000000","probiotic paste"="#FFC080","probiotic bag"="#FF8000","control paste"="#90BFF9","control bag"="#0000C0")
DAplot$Treatment<-factor(DAplot$Treatment,levels=c("probiotic bag","probiotic paste","control bag","control paste","none"))

pdf("DA_Oct.pdf",bg ="white",width=11)
p2<-ggplot(DAplot,aes(x=Treatment,y=proportion,color=Treatment))+
  geom_boxplot()+
  geom_point()+
  facet_grid(.~Genus)+
  theme_bw()+
  scale_y_break(c(0.03, 0.35),scales="free")+
  theme(strip.text.x = element_text(face="italic"))+
  scale_color_manual(values=cols)
p2
dev.off()
# the pdf that has a plot using the ggbreak feature adds a blank page in front of the figure.

```



### Use corncob to look for differentially abundant FAMILIES by category: treatment type

####### OCTOBER
# use an ASV table that has been filtered to remove low-abundance ASVs, no blanks, replaced sequences with ASV numbers in both otu table and taxa table, added a column named "Species" to the taxa table to fill in later with ASV number


```{r, echo=FALSE}
otu <- read.table("silva_nochloronomito_otu_table_ps5_BS2_oct_corncob.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_nochloronomito_taxa_table_ps5_BS2_oct_corncob.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata_ps5_BS2_oct.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps5_BS2_oct <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
                        sample_data(samples), 
                        tax_table(taxon))
ps5_BS2_oct  #756 taxa and 80 samples

# collapse ASVs into higher taxonomic groups for DA analysis
oct_family <- ps5_BS2_oct %>% phyloseq::tax_glom("Family")
oct_family # 183 taxa (families) and 80 samples


# What microorganisms are significantly different by treatment type?

# DA determined relative to first reference level, so set this manually to the "none" treatment
sample_data(oct_family)$Treatment<-factor(sample_data(oct_family)$Treatment,levels=c("none","probiotic bag","probiotic paste","control bag","control paste"))

set.seed(1)
treatment.da <- differentialTest(formula = ~ Treatment, 
                                 phi.formula = ~ Treatment,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Treatment,
                                 test = "Wald", boot = FALSE,
                                 data = oct_family,
                                 fdr_cutoff = 0.05)

summary(treatment.da)
treatment.da
treatment.da$significant_taxa # 3 differentially abundant families

#quick look at results with corncob's plotting feature
plot(treatment.da, c("Family"))
# Positive values mean that taxon is enriched in the treatment in the header relative to no treatment
# DA families: Fusibacteraceae, Rubritaleaceae, and Corynebacteriaceae

# create a function that will return the values of interest

sigtaxto_df <- function(treatment.da, oct_family) {
  df.out <- c()
  for (i in 1:length(treatment.da$significant_models)) { #from 1 to the number of significant taxa
    df.out = rbind(df.out, treatment.da$significant_models[[i]]$coefficients[2, 1:4]) #bind the significant taxa with coefficient, std. error, t value, Pr. coefficients[2,] is the DA model output.
  }
  asv = as.vector(treatment.da$significant_taxa) #get significant families
  Class = as.vector(otu_to_taxonomy(data = oct_family, treatment.da$significant_taxa, level = c("Class")))
  Order = as.vector(otu_to_taxonomy(data = oct_family, treatment.da$significant_taxa, level = c("Order")))
  Family = as.vector(otu_to_taxonomy(data = oct_family, treatment.da$significant_taxa, level = c("Family")))
  
  df.bind <- as.data.frame(cbind(df.out, asv, Class, Order, Family))
  colnames(df.bind) <- c("Estimate", "StdError", "tvalue", "Pr", "asv", "Class", "Order", "Family")
  df.bind$Estimate = as.numeric(as.character(df.bind$Estimate)) #estimate column should be numeric 
  df.bind$StdError = as.numeric(as.character(df.bind$StdError)) #stderror column should be numeric 
  
  return(df.bind)
}

df <- sigtaxto_df(treatment.da, oct_family) #make a df of the corncob output and the taxonomy

### Save these data so you don't have to re-run the model ###
write.table(df, "DA-Families_treatment-Oct.txt", sep="\t",row.names = FALSE)


# retrieve relative abundances of significant taxa for plotting
oct_family #183 taxa and 80 samples
oct_family_ra<-transform_sample_counts(oct_family, function(OTU) OTU/sum(OTU))
oct_family_ra #183 taxa and 80 samples
otu_da = as(otu_table(oct_family_ra), "matrix")
taxon_da = as(tax_table(oct_family_ra), "matrix")
da<-t(otu_da)
da <- as.data.frame(da)
da <- tibble::rownames_to_column(da, "ASV")
taxon_da <- as.data.frame(taxon_da)
taxon_da <- tibble::rownames_to_column(taxon_da, "ASV")
taxon_da$Species<-NULL
taxon_da$Genus<-NULL
da_tax <- merge(taxon_da, da, by="ASV", all=FALSE)

# now to retrieve only the taxa that I want to plot
DA1 <- da_tax %>% filter(Family =='Fusibacteraceae')
DA2 <- da_tax %>% filter(Family =='Rubritaleaceae')
DA3 <- da_tax %>% filter(Family =='Corynebacteriaceae')
DA_oct <-rbind(DA1,DA2,DA3)

# Merge with metadata for plotting aesthetics
samples <- as.data.frame(samples)
samples <- tibble::rownames_to_column(samples, "sample")
DA_long<-melt(DA_oct,value.name="proportion",variable.name="sample",id.vars=c("ASV","Kingdom","Phylum","Class","Order","Family"))

DAplot <- merge(DA_long,samples,by="sample", all=TRUE)
# reduce decimal places in proporation column
DAplot$proportion<-round(DAplot$proportion,3)

cols<-c("none"="#000000","probiotic paste"="#FFC080","probiotic bag"="#FF8000","control paste"="#90BFF9","control bag"="#0000C0")
DAplot$Treatment<-factor(DAplot$Treatment,levels=c("probiotic bag","probiotic paste","control bag","control paste","none"))

pdf("DA-Families_Oct.pdf",bg ="white",width=8)
p2<-ggplot(DAplot,aes(x=Treatment,y=proportion,color=Treatment))+
  geom_boxplot()+
  geom_point()+
  facet_grid(Family~.)+
  theme_bw()+
  scale_color_manual(values=cols)
p2
dev.off()


```

### DISEASE LESION SAMPLES ONLY 

### Use corncob to look for differentially abundant FAMILIES by category: treatment type ONLY IN DISEASE LESION SAMPLES


```{r, echo=FALSE}

### Prepare dataset first
otu <- read.table("silva_nochloronomito_otu_table_ps5_BS2.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_nochloronomito_taxa_table_ps5_BS2.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata_ps5_BS2.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps5_BS2 <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
               sample_data(samples), 
               tax_table(taxon))
ps5_BS2 #756 taxa and 194 samples

#Look at diseased samples only
ps5_BS2_DD = subset_samples(ps5_BS2, Condition == "DD")
ps5_BS2_DD #756 taxa and 79 samples
# also remove August samples (pre-treatment)
ps5_BS2_DD_after = subset_samples(ps5_BS2_DD, Collection.month != "Aug")
ps5_BS2_DD_after #756 taxa and 55 samples
ps5_BS2_DD_after <- prune_taxa(taxa_sums(ps5_BS2_DD_after) > 1, ps5_BS2_DD_after)
ps5_BS2_DD_after #662 taxa and 55 samples

# export reduced tables and replace ASV sequence with ASV#
otu_after = as(otu_table(ps5_BS2_DD_after), "matrix")
taxon_after = as(tax_table(ps5_BS2_DD_after), "matrix")
metadata_after = as(sample_data(ps5_BS2_DD_after), "matrix")
write.table(otu_after,"silva_nochloronomito_otu_table_ps5_BS2_DD_after_corncob.txt",sep="\t",col.names=NA)
write.table(taxon_after,"silva_nochloronomito_taxa_table_ps5_BS2_DD_after_corncob.txt",sep="\t",col.names=NA)
write.table(metadata_after,"metadata_ps5_BS2_DD_after_corncob.txt",sep="\t",col.names=NA)


### Clear the environment and read back in prepped dataset with ASV#s
otu <- read.table("silva_nochloronomito_otu_table_ps5_BS2_DD_after_corncob.txt",sep="\t",header=TRUE, row.names=1)
taxon <- read.table("silva_nochloronomito_taxa_table_ps5_BS2_DD_after_corncob.txt",sep="\t",header=TRUE,row.names=1)
samples<-read.table("metadata_ps5_BS2_DD_after_corncob.txt",sep="\t",header=T,row.names=1)
OTU = otu_table(otu, taxa_are_rows=FALSE)
taxon<-as.matrix(taxon)
TAX = tax_table(taxon)
sampledata = sample_data(samples)
ps5_BS2_DD_after <- phyloseq(otu_table(otu, taxa_are_rows=FALSE), 
                        sample_data(samples), 
                        tax_table(taxon))
ps5_BS2_DD_after #662 taxa and 55 samples


# What microorganisms are significantly different by treatment type?

# DA determined relative to first reference level, so set this manually to the "none" treatment
sample_data(ps5_BS2_DD_after)$Treatment<-factor(sample_data(ps5_BS2_DD_after)$Treatment,levels=c("none","probiotic bag","probiotic paste","control bag","control paste"))

set.seed(1)
treatment.da <- differentialTest(formula = ~ Treatment, 
                                 phi.formula = ~ Treatment,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Treatment,
                                 test = "Wald", boot = FALSE,
                                 data = ps5_BS2_DD_after,
                                 fdr_cutoff = 0.05)

summary(treatment.da)
treatment.da
treatment.da$significant_taxa # 2 differentially abundant ASVs
# ASV20, ASV135

#quick look at results with corncob's plotting feature
plot(treatment.da, c("Family","Genus"))
# Positive values mean that taxon is enriched in the treatment in the header relative to no treatment

# create a function that will return the values of interest
sigtaxto_df <- function(treatment.da, ps5_BS2_DD_after) {
  df.out <- c()
  for (i in 1:length(treatment.da$significant_models)) { #from 1 to the number of significant taxa
    df.out = rbind(df.out, treatment.da$significant_models[[i]]$coefficients[2, 1:4]) #bind the significant taxa with coefficient, std. error, t value, Pr. coefficients[2,] is the DA model output.
  }
  asv = as.vector(treatment.da$significant_taxa) #get significant ASVs
  Class = as.vector(otu_to_taxonomy(data = ps5_BS2_DD_after, treatment.da$significant_taxa, level = c("Class")))
  Order = as.vector(otu_to_taxonomy(data = ps5_BS2_DD_after, treatment.da$significant_taxa, level = c("Order")))
  Family = as.vector(otu_to_taxonomy(data = ps5_BS2_DD_after, treatment.da$significant_taxa, level = c("Family")))
  Genus = as.vector(otu_to_taxonomy(data = ps5_BS2_DD_after, treatment.da$significant_taxa, level = c("Genus")))
  
  df.bind <- as.data.frame(cbind(df.out, asv, Class, Order, Family, Genus))
  df.bind$asvgenus <- paste0(df.bind$Genus, " (", df.bind$asv, ")")
  colnames(df.bind) <- c("Estimate", "StdError", "tvalue", "Pr", "asv", "Class", "Order", "Family", "Genus", "asvtaxa")
  df.bind$Estimate = as.numeric(as.character(df.bind$Estimate)) #estimate column should be numeric 
  df.bind$StdError = as.numeric(as.character(df.bind$StdError)) #stderror column should be numeric 
  
  return(df.bind)
}

df <- sigtaxto_df(treatment.da, ps5_BS2_DD_after) #make a df of the corncob output and the taxonomy

### Save these data so you don't have to re-run the model ###
write.table(df, "DA_treatment-DD_after.txt", sep="\t",row.names = FALSE)

# retrieve relative abundances of significant taxa for plotting
ps5_BS2_DD_after #662 taxa and 55 samples
ps5_BS2_DD_after_ra<-transform_sample_counts(ps5_BS2_DD_after, function(OTU) OTU/sum(OTU))
ps5_BS2_DD_after_ra #662 taxa and 55 samples
otu_da = as(otu_table(ps5_BS2_DD_after_ra), "matrix")
taxon_da = as(tax_table(ps5_BS2_DD_after_ra), "matrix")
da<-t(otu_da)
da <- as.data.frame(da)
da <- tibble::rownames_to_column(da, "ASV")
taxon_da <- as.data.frame(taxon_da)
taxon_da <- tibble::rownames_to_column(taxon_da, "ASV")
taxon_da$Species<-NULL
da_tax <- merge(taxon_da, da, by="ASV", all=FALSE)

# now to retrieve only the taxa that I want to plot
DA1 <- da_tax %>% filter(ASV =='ASV20')
DA2 <- da_tax %>% filter(ASV =='ASV135')
DA_DD_after <-rbind(DA1,DA2)

# Merge with metadata for plotting aesthetics
samples <- as.data.frame(samples)
samples <- tibble::rownames_to_column(samples, "sample")
DA_long<-melt(DA_DD_after,value.name="proportion",variable.name="sample",id.vars=c("ASV","Kingdom","Phylum","Class","Order","Family","Genus"))

DAplot <- merge(DA_long,samples,by="sample", all=TRUE)
# reduce decimal places in proporation column
DAplot$proportion<-round(DAplot$proportion,3)

cols<-c("none"="#000000","probiotic paste"="#FFC080","probiotic bag"="#FF8000","control paste"="#90BFF9","control bag"="#0000C0")
DAplot$Treatment<-factor(DAplot$Treatment,levels=c("probiotic bag","probiotic paste","control bag","control paste","none"))

pdf("DA_DD_after.pdf",bg ="white",width=8)
p2<-ggplot(DAplot,aes(x=Treatment,y=proportion,color=Treatment))+
  geom_boxplot()+
  geom_point()+
  facet_grid(.~Genus)+
  theme_bw()+
  scale_y_break(c(0.01, 0.5),scales="free")+
  theme(strip.text.x = element_text(face="italic"))+
  scale_color_manual(values=cols)
p2
dev.off()



###### DA families

# collapse ASVs into higher taxonomic groups for DA analysis
DD_family <- ps5_BS2_DD_after %>% phyloseq::tax_glom("Family")
DD_family # 161 taxa (families) and 55 samples

# DA determined relative to first reference level, so set this manually to the "none" treatment
sample_data(DD_family)$Treatment<-factor(sample_data(DD_family)$Treatment,levels=c("none","probiotic bag","probiotic paste","control bag","control paste"))

set.seed(1)
treatment.da <- differentialTest(formula = ~ Treatment, 
                                 phi.formula = ~ Treatment,
                                 formula_null = ~ 1,
                                 phi.formula_null = ~ Treatment,
                                 test = "Wald", boot = FALSE,
                                 data = DD_family,
                                 fdr_cutoff = 0.05)

summary(treatment.da)
treatment.da
treatment.da$significant_taxa # 2 differentially abundant families

#quick look at results with corncob's plotting feature
plot(treatment.da, c("Family"))
# Positive values mean that taxon is enriched in the treatment in the header relative to no treatment
# DA families: Peptostreptococcaceae, family classified only to "Bacteria"


# create a function that will return the values of interest
sigtaxto_df <- function(treatment.da, DD_family) {
  df.out <- c()
  for (i in 1:length(treatment.da$significant_models)) { #from 1 to the number of significant taxa
    df.out = rbind(df.out, treatment.da$significant_models[[i]]$coefficients[2, 1:4]) #bind the significant taxa with coefficient, std. error, t value, Pr. coefficients[2,] is the DA model output.
  }
  asv = as.vector(treatment.da$significant_taxa) #get significant families
  Class = as.vector(otu_to_taxonomy(data = DD_family, treatment.da$significant_taxa, level = c("Class")))
  Order = as.vector(otu_to_taxonomy(data = DD_family, treatment.da$significant_taxa, level = c("Order")))
  Family = as.vector(otu_to_taxonomy(data = DD_family, treatment.da$significant_taxa, level = c("Family")))
  
  df.bind <- as.data.frame(cbind(df.out, asv, Class, Order, Family))
  colnames(df.bind) <- c("Estimate", "StdError", "tvalue", "Pr", "asv", "Class", "Order", "Family")
  df.bind$Estimate = as.numeric(as.character(df.bind$Estimate)) #estimate column should be numeric 
  df.bind$StdError = as.numeric(as.character(df.bind$StdError)) #stderror column should be numeric 
  
  return(df.bind)
}

df <- sigtaxto_df(treatment.da, DD_family) #make a df of the corncob output and the taxonomy

### Save these data so you don't have to re-run the model ###
write.table(df, "DA-Families_treatment-DD_after.txt", sep="\t",row.names = FALSE)


# retrieve relative abundances of significant taxa for plotting
DD_family #161 families and 55 samples
DD_family_ra<-transform_sample_counts(DD_family, function(OTU) OTU/sum(OTU))
DD_family_ra #161 families and 55 samples
otu_da = as(otu_table(DD_family_ra), "matrix")
taxon_da = as(tax_table(DD_family_ra), "matrix")
da<-t(otu_da)
da <- as.data.frame(da)
da <- tibble::rownames_to_column(da, "ASV")
taxon_da <- as.data.frame(taxon_da)
taxon_da <- tibble::rownames_to_column(taxon_da, "ASV")
taxon_da$Species<-NULL
taxon_da$Genus<-NULL
da_tax <- merge(taxon_da, da, by="ASV", all=FALSE)

# now to retrieve only the taxa that I want to plot
DA1 <- da_tax %>% filter(Family =='Peptostreptococcaceae')
DA2 <- da_tax %>% filter(Family =='Bacteria')
DA_DD <-rbind(DA1,DA2)

# Merge with metadata for plotting aesthetics
samples <- as.data.frame(samples)
samples <- tibble::rownames_to_column(samples, "sample")
DA_long<-melt(DA_DD,value.name="proportion",variable.name="sample",id.vars=c("ASV","Kingdom","Phylum","Class","Order","Family"))

DAplot <- merge(DA_long,samples,by="sample", all=TRUE)
# reduce decimal places in proporation column
DAplot$proportion<-round(DAplot$proportion,3)

cols<-c("none"="#000000","probiotic paste"="#FFC080","probiotic bag"="#FF8000","control paste"="#90BFF9","control bag"="#0000C0")
DAplot$Treatment<-factor(DAplot$Treatment,levels=c("probiotic bag","probiotic paste","control bag","control paste","none"))

pdf("DA-Families_Oct.pdf",bg ="white",width=8)
p2<-ggplot(DAplot,aes(x=Treatment,y=proportion,color=Treatment))+
  geom_boxplot()+
  geom_point()+
  facet_grid(Family~.,scales="free")+
  theme_bw()+
  scale_color_manual(values=cols)
p2
dev.off()



```


